Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-04-26 Paper/paper.tex
changed method & citation

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban 
et al. \cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}; these 
are instances in which two highly similar sequences fold into remarkably different structures. Orban and co-workers have designed a 
protein-G-like sequence of 56-residues that
is marginally stable in one of two possible folds. By mutating key residues in this sequence they
are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We call sequences ``$\beta$'' that fold into 
the protein-G-like that fold into a $4 \beta + \alpha$ fold, and we call sequences ``$\alpha$'' that fold into $3 \alpha$ helical folds.  One pair 
of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair (GA95/GB95) is 95 percent identical, 
differing in three positions. The third pair (GA98/GB98) differs in only a single tyrosine-to-alanine mutation. Accurately predicting the 
structural preferences of these structures presents a serious challenge for computational methods 
\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}.

@@

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all four sequences (data not shown). This is, however, not
surprising. It is well known [CITE]\Arijit{still searching this citation} that it is easy to distinguish 
computational models from native structures. To avoid this potential problem, we instead computed relative free energy of two
different computer generated models for each sequence. One model is based on the $\alpha$ structure and
the other on the $\beta$ structure (see Supporting Information for details on the modeling procedure).
This is a much more realistic test of the confinement method's ability to accurately calculate
relative free energies.

@@

\Ken{[If this is a relevant limitation, let's say something about it.]}
\Arijit{I think Ken is pointing towards the normal mode issue and number of residues that we can handle}

@@

All calculations where performed with the amber 11 suit of programs in combination with ff99SB and
GB/SA implicit solvent. Interestingly, we extend the method for calculation of per residue free
energy in an approximate way. For this purpose, the confinement energy, $\Delta G_{A,A*}$ and
$\Delta G_{B,B*}$ of each residue is calculated in the usual numerical way as described above. The
internal energy of each residue is calculated using the decomp module of amber from the final
restrained trajectory. We call this method approximate as we ignore the contribution from the normal
mode or quasiharmonic analysis. However, this contrbution to the total free energy is much smaller, 
which allow us to study the mechanistic details of conformational preference of each residue.

@@

@@

>>>>

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban 
et al. \cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}; these 
are instances in which two highly similar sequences fold into remarkably different structures. Orban and co-workers have designed a 
protein-G-like sequence of 56-residues that
is marginally stable in one of two possible folds. By mutating key residues in this sequence they
are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We call conformation ``$\beta$'' that fold into 
the protein-G-like that fold into a $4 \beta + \alpha$ fold, and we call conformation ``$\alpha$'' that fold into $3 \alpha$ helical folds.  One pair 
of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair (GA95/GB95) is 95 percent identical, 
differing in three positions. 
%The third pair (GA98/GB98) differs in only a single tyrosine-to-alanine mutation. 
Accurately predicting the 
structural preferences of these structures presents a serious challenge for computational methods 
\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}.

@@

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all four sequences (data not shown). This is, however, not
surprising. It is well known \cite{Handl2009},\cite{Sheffler2009} that it is easy to distinguish 
computational models from native structures. To avoid this potential problem, we instead computed relative free energy of two
different computer generated models for each sequence. One model is based on the $\alpha$ structure and
the other on the $\beta$ structure (see Supporting Information for details on the modeling procedure).
This is a much more realistic test of the confinement method's ability to accurately calculate
relative free energies.

@@

\Ken{[If this is a relevant limitation, let's say something about it.]}
\Arijit{I think we need to write the normal mode issue and number of residues that we can handle. If 
one of you contribute here that will be great.}

@@

All calculations where performed with the amber 11 suit of programs \cite{Case2012},\cite{Goetz2012} in
combination with ff99SB forcefield\cite{Hornak2006} and generalized born implicit solvent \cite{Mongan2006}.
Interestingly, we extend the method for calculation of per residue free
energy in an approximate way. For this purpose, the confinement energy, $\Delta G_{A,A*}$ and
$\Delta G_{B,B*}$ of each residue is calculated in the usual numerical way as described in ref. by Tyka 
et al. \cite{Tyka2006}. The internal energy of each residue is calculated using the decomp module of amber from the final
restrained trajectory. We call this method approximate as we ignore the entropic contribution from the normal
mode or quasiharmonic analysis. However, this contrbution to the total free energy is much smaller,
which allow us to study the mechanistic details of conformational preference of each residue.

@@

\bibitem{Handl2009}
Handl, J.; Knowles, J.; Lovel, S.C. Artefacts and biases affecting the evaluation of scoring
functions on decoy sets for protein structure prediction. Bioinformatics, 2009, 25, 1271-1279.

@@

\bibitem{Hornak2006}
Hornak, V.; Abel, R.; Okur, A.; Strockbine, B.; Roitberg, A.; Simmerling, C. Comparison of multiple Amber force fields
and development of improved protein backbone parameters. Proteins., 2006, 65, 712-725.

\bibitem{Mongan2006}
Mongan, J.; Simmerling, C.; A. McCammon, J.; A. Case, D.; Onufriev, A. Generalized
Born with a simple, robust molecular volume correction. J. Chem. Theory Comput.,
2006, 3, 156-169.
